---
title: "PS 6"
author: "Arvind Patel arvindp@mtu.edu"
date: "October, 2018"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 1: Categorical relationships

Consider the following data sets, which are the results of a survey where children were asked two questions. First, they were asked "At an indoor birthday party, would you prefer a. clowns; b. horses; c. chocolate fountain; d. magician?".  Then they were asked the same thing about outdoor parties.  tab1 looks at the cross-tabulations of each pair of responses.  tab2 looks at hypothetical marginal values--what responses were given for indoor and outdoor across the four options.

```{r}
set.seed(1000)
indoor<- sample(c("A","B","C","D"),prob=c(5,8,25,6),
   size=200,replace=T)
outdoor <-sample(c("A","B","C","D"),prob=c(5,20,3,6),
   size=200,replace=T)

tab1 <- table(indoor,outdoor)
tab2 <- table(c(indoor,outdoor),rep(c("I","O"),each=200))
tab1
tab2

```

Conduct both chi-squared tests and bayes factor contingency table tests for tab1 and tab2 to determine (1) whether responses to indoor preference influenced individuals response to outdoor preference, and (2) to determine whether indoor and outdoor preferences were the same.  Describe the meaning of the result, and interpret specifically what it means for the particular table. Your explanation should differ for the two tables, regardless of whether they are significant or not.

```{r}
chisq.test(tab1)
```
Null Hypothesis: 
There is not a significant association between indoor and outdoor variables, the variables are independent of each other; any association between them is likely due to chance and sampling error.
Alternative Hypothesis:
There is a significant (positive or negative) association between variables; the association between variables is likely not due to chance or sampling error

Since the p value =0.827 >0.05, we reject the null hypothesis and conclude that there IS a significant association between the indoor and outdoor variables.
```{r}
chisq.test(tab2)
```
Null Hypothesis: 
There is not a significant association between indoor and outdoor variables, the variables are independent of each other; any association between them is likely due to chance and sampling error.
Alternative Hypothesis:
There is a significant (positive or negative) association between variables; the association between variables is likely not due to chance or sampling error

Since the p value < 0.05, we accept the null hypothesis and conclude that there is NO significant association between the indoor and outdoor variables.

```{r}
library(BayesFactor)
contingencyTableBF(tab1, sampleType = "jointMulti")
contingencyTableBF(tab2, sampleType = "jointMulti")
```
Null Hypothesis: 
There is not a significant association between indoor and outdoor variables, the variables are independent of each other; any association between them is likely due to chance and sampling error.
Alternative Hypothesis:
There is a significant (positive or negative) association between variables; the association between variables is likely not due to chance or sampling error


For table 1, when Bayes factor analyis is performed we get Bayes factor value < 1/100 which implies Extreme evidence for the H0

For table 2, when Bayes factor analyis is performed we get Bayes factor value >100 which implies Extreme evidence for the H1





#Problem 2: Which test to do: You decide.

For the following example data sets and questions, answer the statistical question using an appropriate test.  Whenever possible, run both a NHST and a Bayesian test.  If you choose to use a non-parametric test, give a rationale for why.  The data here are completely fabricated, so do not use your intuition for what should be true, but rather find out from the data.

The column names are:
 * age: age of car owner
 * gender: gender of car owner
 * type: type of vehicle
 * origin: location car was manufactured
 * origin.last: location of previous car's manufacture
 * carval: purchase price of vehicle
 * carval.last purchase price of previous vehicle
 
 

```{r}

cardat <- read.table(text="age gender  type origin origin.last carval carval.last
   34      F   SUV     US          US  16400       15800
   31      M Truck     US      Europe  16900       16000
   47      M Sedan     US          US  18800       17100
   21      F Sedan  Japan       Japan  16000       15500
   42      M   SUV     US       Japan  16800       16100
   43      F   SUV     US          US  17200       16300
   60      F Truck Europe      Europe  19900       17800
   37      M Truck Europe      Europe  17100       16200
   46      F   SUV  Japan       Japan  16900       16300
  27      M Sedan     US          US  16200       15700
  50      M   SUV     US          US  18800       17100
  64      F   SUV  Japan          US  50700       31700
  33      M   SUV  Japan       Japan  16500       15900
  39      M Truck     US      Europe  17000       16200
  58      F Sedan  Japan          US  19400       17500
  53      F   SUV     US      Europe  19200       17400
  29      F Sedan     US       Japan  16300       15700
  37      F Sedan     US          US  17300       16300
  37      M   SUV     US       Japan  18200       16700
  54      F Sedan  Japan       Japan  24500       19800
  46      F   SUV  Japan      Europe  18000       16700
  55      F   SUV     US       Japan  28900       21700
  46      F Truck     US      Europe  16600       16100
  57      M   SUV Europe      Europe  24300       19700
  40      M   SUV     US          US  16800       16100
  27      M Sedan  Japan          US  16900       16000
  58      M   SUV Europe      Europe  20300       17900
  64      M Truck     US          US  40600       27100
  47      M Truck     US      Europe  18400       16900
  32      M Truck     US          US  15900       15600
  43      F Sedan  Japan          US  17200       16300
  66      M Truck Europe      Europe  19100       17500
  36      F   SUV     US       Japan  16900       16100
  68      M Truck     US          US  69300       40100
  54      F Sedan  Japan          US  17000       16400
  64      M Truck  Japan      Europe  34900       24600
  27      M   SUV  Japan      Europe  15800       15500
  51      F Sedan  Japan       Japan  29000       21700
  69      M Sedan     US       Japan  54400       33400
  25      F Sedan  Japan       Japan  15800       15500",header=T)
cardat
```
Answer each of the following:
* Is there an impact of gender on the type of car purchased?
```{r}
library(MASS)
tabl <- table(cardat$gender,cardat$type)
tabl
chisq.test(tabl)
plot(cardat$gender,cardat$type, xlab = "Gender",ylab = "Type of Car", col= c("orange", "light blue","grey"),main = "Gender by type of car")
```
Null Hypothesis: 
There is not a significant association between gender and type of the car, the variables are independent of each other; any association between them is likely due to chance and sampling error.
Alternative Hypothesis:
There is a significant (positive or negative) association between gender and type of the car; the association between variables is likely not due to chance or sampling error

Since the p value < 0.05, we accept the null hypothesis and conclude that there is NO significant association between the gender and type of car.

From the chart above, we find that Females buy more Sedan than males and Males buy more trucks than females but since the statistical significance test performed states that it may have happend just by chance and there is not real association in gender and type of car.



* Is there a difference in amount paid for a car for men versus women?
```{r}
library(ggplot2)
ggplot(cardat, aes(cardat$gender, cardat$carval)) +
        geom_boxplot(colour = "blue") + labs(title = "Gender v/s Car value") +  theme(axis.text.x=element_text(color = "black", size=11)) + ylab("Car Value") + xlab("Gender")
t.test(cardat$carval~cardat$gender, data = cardat)

```
The null hypothesis is that the two means are equal, and the alternative is that they are not.
We observe that, the p value is not less than 0.05, hence the results are not significant and we reject the null hypothesis and conclude that the means are not equal

From the boxplot we observe that, the median value of car value for males is more than females and the statistical test too support the statement. Mean value of car value for male is more than the females.


* Do people tend to buy vehicles from of the same origin as their last vehicle (US, europe, japan)?
```{r}
tbl <- table(cardat$origin, cardat$origin.last)
chisq.test(tbl)
plot(cardat$origin.last,cardat$origin, ylab = "Origin",xlab = "Last Origin", col= c("orange", "light blue","grey"),main = "Comparision of Origin by last origin")
```
Null Hypothesis: 
There is not a significant association between origin and last origin of the car, the variables are independent of each other; any association between them is likely due to chance and sampling error.
Alternative Hypothesis:
There is a significant (positive or negative) association between origin and last origin of the car; the association between variables is likely not due to chance or sampling error

Since the p value < 0.05, we accept the null hypothesis and conclude that there is NO significant association between the origin and last origin of the car.

From the plot, we can see that if the last origin was Japan or US they retained to either Japan or US and didn't switch to European cars. People who had last origin as US had retained to US cars in major proportion.


* Is there a relationship between driver age and the value of he car?
```{r}
t.test(cardat$age,cardat$carval)
linear_model<-lm(cardat$carval~cardat$age)
plot(cardat$age, cardat$carval, xlab = "Age", ylab = "Car value", main = "Car value by age")
abline(linear_model$coeff,lwd=3,col="red")
```
The null hypothesis is that the mean age value and mean car value are equal, and the alternative is that they are not.
We observe that, the p value is less than 0.05, hence the results are significant and we do not reject the null hypothesis and conclude that the mean age value and mean car value are equal.
From the plot above, we see that age and car value follows a trend and it is given by the regression line. As the age increases, we find the car values increases too



* What is your best estimate for the value of a car driven by a 32, 52, and 62-year-old?

```{r}

linear_model<-lm(cardat$carval~cardat$age)
linear_model$coefficients
fitted(linear_model)
plot(cardat$age, cardat$carval, xlab = "Age", ylab = "Car value", main = "Car value by age")
abline(linear_model$coeff,lwd=3,col="red")

```
From the table of fitted values and fitted linear regression line, we find the best estimated values. 


* Is there a relationship between how much someone paid for their previous car and how much they paid for their current car?
* Did people tend to pay more for their current car than their previous car?

```{r}
t.test(cardat$carval,cardat$carval.last , paired = TRUE)
plot(cardat$carval,cardat$carval.last, xlab = "Car value", ylab = "Last Car value", main = "Car value compared to last car value")
```
The null hypothesis (H0) assumes that the true mean difference (??d) is equal to zero.
The alternative hypothesis (H1) assumes that ??d is not equal to zero

we reject H0 (because p<0.05) and state that we have significant evidence that car values changed from changed from last car values.

From the plot above, we see that there is linear trend in the car values from the last car values and also we find that car values have changed significantly from the last car values and they car value is almost double the last car values in some cases.


* Did trucks cost more than SUVs?
```{r}
type_truck <- cardat[cardat$type == 'Truck',]
type_suv <- cardat[cardat$type == 'SUV',]
t.test(type_truck$carval,type_suv$carval)
boxplot(type_truck$carval,type_suv$carval, xlab = "Truck and SUVs", ylab = "Cost", main = "Cost of Truck and SUVs", col = c("red", "light blue"))
axis(side=1,at=c(1,2),labels=c("Truck","SUV"))
```
The null hypothesis is that the mean cost of truck and SUVs are equal, and the alternative hypothesis is that mean cost of truck and SUVs are not equal.

We observe that, the p value is not less than 0.05, hence the results are not significant and we reject the null hypothesis and conclude that the mean cost of truck and SUVs are not equal and infact Truck cost more than SUVs.

From the box plot, we find that the median cost of truck was greater than SUVs. Also it had lot of variation in cost as compared to SUV
