
```{r}
library(ggplot2)
library(caret)
library(AppliedPredictiveModeling)
#read the music genre data
genre_data <-read.csv('genresTrain.csv')
#check the dimensions
dim(genre_data)
str(genre_data)
#frequency distribution
ggplot(genre_data, aes(GENRE)) + geom_bar()

#question 4.1(b) and (c)
set.seed(111)

#partition the data using stratified sampling
train_d <- createDataPartition(genre_data$GENRE, p = .80, list= FALSE)

#finding training data
train_data <-genre_data[train_d,]

#Finding testing data
test_data <-genre_data[-train_d,]

#K-fold cross validation
set.seed(111)
k_folds <- createFolds(train_d,k=10, list = FALSE, returnTrain = TRUE)
train_cv <- genre_data[k_folds[[1]],]

```

```{r}
data(permeability)
ggplot(as.data.frame(permeability), aes(permeability)) + geom_histogram(color= "black",fill = "grey", bins = 5)
```

```{r}
#question 4.2 (b) and (c)
set.seed(111)
p_folds <- createMultiFolds(permeability,k=3)
p_cv <- permeability[p_folds[[1]]]
f_cv <- fingerprints[p_folds[[1]]]
```
```{r}
#question 4.3(b)
comp=c(1:10)
mu=c(0.444,0.500,0.533,0.545,0.542,0.537,0.534,0.534,0.520,0.507)
Std_error=c(0.0272,0.0298,0.0302,0.0308,0.0322,0.0327,0.0333,0.0330,0.0326,0.0324)
maxi=max(Mean)
tolerance=round(((mu-maxi)/maxi)*100,10)
tolerance_data=cbind(comp,mu,Std_error,tolerance)
colnames(tolerance_data)=c('Components','Mean','Standard Error','Tolerance percent')
tolerance_data
```

```{r}
#Question 4.4

#read the oil data
set.seed(111)
data(oil)
str(oilType)
oiltable=table(oilType)
print(oiltable/length(oilType))

#part(A)
set.seed(111)
oil_s=matrix(rep(0,60*10),nrow=60)
for (i in 1:10)
{
  oil_s[,i]=sample(oilType,60,replace=TRUE)
}
table(oil_s[,1])/length(oil_s[,1])
table(oil_s[,2])/length(oil_s[,2])
table(oil_s[,3])/length(oil_s[,3])
table(oil_s[,4])/length(oil_s[,4])
table(oil_s[,5])/length(oil_s[,5])
table(oil_s[,6])/length(oil_s[,6])
table(oil_s[,7])/length(oil_s[,7])
table(oil_s[,8])/length(oil_s[,8])
table(oil_s[,9])/length(oil_s[,9])
table(oil_s[,10])/length(oil_s[,10])


#part(B)
stratified_sampled=createDataPartition(oilType,p=0.60,times=7)
table(oilType[stratified_sampled$Resample1])/length(stratified_sampled$Resample1)
table(oilType[stratified_sampled$Resample2])/length(stratified_sampled$Resample2)
table(oilType[stratified_sampled$Resample3])/length(stratified_sampled$Resample3)
table(oilType[stratified_sampled$Resample4])/length(stratified_sampled$Resample4)
table(oilType[stratified_sampled$Resample5])/length(stratified_sampled$Resample5)
table(oilType[stratified_sampled$Resample6])/length(stratified_sampled$Resample6)
table(oilType[stratified_sampled$Resample7])/length(stratified_sampled$Resample7)



